---
title: "Lab 6 Homework"
author: "Richard J. Dean"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(skimr)
```

For this assignment we are going to work with a large data set from the [United Nations Food and Agriculture Organization](http://www.fao.org/about/en/) on world fisheries. These data are pretty wild, so we need to do some cleaning. First, load the data.  

Load the data `FAO_1950to2012_111914.csv` as a new object titled `fisheries`.
```{r}
fisheries<-readr::read_csv("data/FAO_1950to2012_111914.csv")
```

1. Do an exploratory analysis of the data (your choice). What are the names of the variables, what are the dimensions, are there any NA's, what are the classes of the variables?  
```{r}
dim(fisheries)
```
```{r}
names(fisheries)
```
```{r}
anyNA(fisheries)
```
```{r}
summary(fisheries)
```
```{r}
t(t(sapply(fisheries,class)))
```
```{r}
skim(fisheries)
```
The names of the variables are: Country, Common name, ISSCAAP group#, ISSCAAP taxonoimic group, ASFIS species#, ASFIS species name, FAO major fishing area, Measure, followed by year numbers 1950-2012. 
The dimensions are 17692x71
There are NA's in the data
The Classes of the variable are 69 Character with 2 Numeric
2. Use `janitor` to rename the columns and make them easier to use. As part of this cleaning step, change `country`, `isscaap_group_number`, `asfis_species_number`, and `fao_major_fishing_area` to data class factor. 
```{r}
fisheries<-janitor::clean_names(fisheries)
```

```{r}
fisheries$country <- as.factor(fisheries$country)
fisheries$isscaap_group_number <- as.factor(fisheries$isscaap_group_number)
fisheries$asfis_species_number <- as.factor(fisheries$asfis_species_number)
fisheries$fao_major_fishing_area <- as.factor(fisheries$fao_major_fishing_area)
```

We need to deal with the years because they are being treated as characters and start with an X. We also have the problem that the column names that are years actually represent data. We haven't discussed tidy data yet, so here is some help. You should run this ugly chunk to transform the data for the rest of the homework. It will only work if you have used janitor to rename the variables in question 2!
```{r}
fisheries_tidy <- fisheries %>% 
  pivot_longer(-c(country,common_name,isscaap_group_number,isscaap_taxonomic_group,asfis_species_number,asfis_species_name,fao_major_fishing_area,measure),
               names_to = "year",
               values_to = "catch",
               values_drop_na = TRUE) %>% 
  mutate(year= as.numeric(str_replace(year, 'x', ''))) %>% 
  mutate(catch= str_replace(catch, c(' F'), replacement = '')) %>% 
  mutate(catch= str_replace(catch, c('...'), replacement = '')) %>% 
  mutate(catch= str_replace(catch, c('-'), replacement = '')) %>% 
  mutate(catch= str_replace(catch, c('0 0'), replacement = ''))

fisheries_tidy$catch <- as.numeric(fisheries_tidy$catch)
fisheries_tidy
```

3. How many countries are represented in the data? Provide a count and list their names.
```{r}
fisheries_tidy %>% 
  summarize(distinct.country=n_distinct(country))
```

```{r}
fisheries_tidy %>% 
  count(country,sort=T)
```
203 countries
4. Refocus the data only to include only: country, isscaap_taxonomic_group, asfis_species_name, asfis_species_number, year, catch.
```{r}
fisheries_tidy %>% 
  select(country, isscaap_taxonomic_group, asfis_species_name, asfis_species_number, year, catch)
```

5. Based on the asfis_species_number, how many distinct fish species were caught as part of these data?
```{r}
fisheries_tidy %>% 
  summarize(distinct.asfis_species_number=n_distinct(asfis_species_number))
```
1551 different species
6. Which country had the largest overall catch in the year 2000?
```{r}
fisheries_tidy %>%
  filter(year=="2000") %>% 
  group_by(country) %>% 
  summarize(sumcatch=sum(catch,na.rm=T),
            totaln=n(),.groups= 'keep') %>% 
  arrange(desc(sumcatch))
```
China had the largest overall catch in 2000
7. Which country caught the most sardines (_Sardina pilchardus_) between the years 1990-2000?
```{r}
fisheries_tidy %>% 
  filter(year>=1990, year<=2000,asfis_species_name=="Sardina pilchardus") %>%
  group_by(country) %>% 
  summarize(catchsum=sum(catch,na.rm=T), .groups= 'keep') %>% 
  arrange(desc(catchsum))
```
Morocco caught the most sardines between 1990 to 2000

8. Which five countries caught the most cephalopods between 2008-2012?
```{r}
fisheries_tidy %>% 
  filter(year>=2008,year<=2012, asfis_species_name=="Cephalopoda") %>% 
  group_by(country) %>% 
  summarize(sumcatch=sum(catch,na.rm=T),.groups= 'keep') %>% 
  arrange(desc(sumcatch))
```
India, China, Spain, Algeria and France
<<<<<<< HEAD
```{r}
summary(fisheries_tidy)
```

9. Which species had the highest catch total between 2008-2012? (hint: Osteichthyes is not a species)
```{r}
fisheries_tidy %>%
  filter(year>=2008, year<=2012) %>%
  group_by(asfis_species_name) %>%
  summarize(species_catch_count=sum(catch,na.rm=T), .groups= 'keep') %>% 
  arrange(desc(species_catch_count))
```
Theragra chalcogramma had the highest catch total between 2008 to 2012
10. Use the data to do at least one analysis of your choice.
```{r}
fisheries_tidy %>% 
  filter(country=="China") %>% 
  group_by(country) %>% 
  summarize(total.catch=sum(catch,na.rm=T),
            std.catch=sd(catch,na.rm=T),
            mean.catch=mean(catch,na.rm=T), .groups= 'keep')
```

## Push your final code to GitHub!
Please be sure that you check the `keep md` file in the knit preferences.   